---
title: "homework_ten"
author: "Isaac Buabeng"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Description
In this homework, I explore the differences in the traits of Palm tree species of African and American/European descent. Specifically, I take look at differences in the average fruit length and maximum stem height asking if there are any differences in these traits. The datasets used for this homework came from [tidytuesday's weekly list of datasets for the year 2025](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-03-18/readme.md).

citation: 
  @misc{tidytuesday, 
    title = {Tidy Tuesday: A weekly social data project}, 
    author = {Data Science Learning Community}, 
    url = {https://tidytues.day}, 
    year = {2024} 
  }



# Variables Used

1. max_stem_height_m 

2. average_fruit_length_cm



# Data Exploration and Cleaning
## load the necessary libraries
```{r}
library(tidyr)
library(ggbeeswarm)
library(cowplot)
library(ggridges)
library(ggmosaic)
library(ggplot2)
library(tidyverse)
library(skimr)
```


## import the data
```{r}
plm_data <- read.csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-18/palmtrees.csv')

str(plm_data)

head(plm_data) 

tail(plm_data)

summary(plm_data)

# check for the different types of tribes we have in our dataset
unique(plm_data$palm_tribe) # 23 tribes in total

# select for species in tribes from africa and america/europe based on (Baker & Dransfield, 2016 and Eiserhardt et al., 2011) 
afro_tribe <- c("Phoeniceae", "Borasseae", "Oranieae", "Sclerospermeae", "Pelagodoxeae", "Podococceae")

eu_us_tribe <- c("Cocoseae", "Phytelepheae", "Geonomateae", "Ceroxyleae", "Chamaedoreeae", "Cryosophileae", "Trachycarpeae", "Irarteeae", "Lepidocaryeae", "Euterpeae", "Eugeissoneae", "Leopoldinieae", "Manicarieae", "Reinhardtieae", "Roystoneeae", "Sabaleae", "Cyclosphatheae")

# select for tribes from africa and take out NAs
plm_data %>%
  filter(palm_tribe %in% afro_tribe) %>%
  drop_na() -> afro_palm 

head(afro_palm)

tail(afro_palm)

# select for tribes from EU and US and take out NAs
plm_data %>% 
  filter(palm_tribe %in% eu_us_tribe) %>%
  drop_na() -> eu_us_palm 

head(eu_us_palm)

# excellent! this would be the only data wrangling we would be doing for now. 
```



# Data Visualization
## visualizing interactions for Tribes in Africa 
```{r}
# mosaic plot for Average Fruit Length by tribe
# Create a binned category for fruit length
afro_palm$fruit_length_cat <- cut(afro_palm$average_fruit_length_cm,
                                  breaks = 4, 
                                  labels = c("Short", 
                                             "Medium", 
                                             "Long", 
                                             "Very Long"))

# Mosaic plot with two categorical variables
ggplot(data = afro_palm) +
  geom_mosaic(aes
              (weight = 1,
                  x = product(palm_tribe), 
                fill = fruit_length_cat)) +
  theme_classic() +
  labs(title = "Fruit Length Categories by Tribe",
       x = "Palm Tribe",
       y = "Proportion",
       fill = "Fruit Length Category")



# Boxplot for Maximum Stem Height by Tribe
ggplot(afro_palm)+ 
  aes(x = palm_tribe, 
      y = max_stem_height_m, 
      fill = palm_tribe) +
  geom_col() +
  theme_cowplot() +
  labs(title = "Maximum Stem Height by Tribe", 
       x = "Tribe", 
       y = "Maximum Stem Height (m)")
```



## visualizing interactions for Tribes in the EU and US
```{r}
# ridgeline plot for Average Fruit Length across tribes
ggplot(data = eu_us_palm,
       aes(
         x    = average_fruit_length_cm,  # continuous
         y    = palm_tribe,               # grouping strip
         fill = palm_tribe                # fill by tribe
       )) +
  geom_density_ridges(alpha = 0.7, color = "white") +
  theme_classic() +
  theme(
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    legend.position = "right"
  ) +
  labs(
    title = "Distribution of Average Fruit Length by Tribe",
    x     = "Average Fruit Length (cm)",
    y     = "Tribe"
  )


# Density ridgelines for Maximum Stem Height across tribes
ggplot(data = eu_us_palm,
       aes(
         y = max_stem_height_m,  # continuous axis
         x = palm_tribe,         # categorical strips
         color = palm_tribe          # map tribe to point color
       )) +
  geom_beeswarm(
    cex       = 3,          # controls horizontal spacing
    size      = 2,          # point size
  ) +
  theme_cowplot() +
  labs(
    title = "Maximum Stem Height by Tribe",
    x     = "Tribe",
    y     = "Maximum Stem Height (m)" 
  ) +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 12, 
                              angle = 45,
                             vjust = 0.5,
                             hjust = 1)
  )

# comparing plots for the African and the American/European tribes it appears that, species in Europe and America have higher averages for Stem Height and Fruit Length. Also, it appears that our data has more records for American/European tribes than thier african counterparts. This is very interesting and needs further investigation.
```

