---
title: "homework_ten"
author: "Isaac Buabeng"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Description
In this homework, I explore the differences in the traits of Palm tree species of African and American/European descent. Specifically, I take look at differences in the average fruit length and maximum stem height asking if there are any differences in these traits. The datasets used for this homework came from [tidytuesday's weekly list of datasets for the year 2025](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-03-18/readme.md).

citation: 
  @misc{tidytuesday, 
    title = {Tidy Tuesday: A weekly social data project}, 
    author = {Data Science Learning Community}, 
    url = {https://tidytues.day}, 
    year = {2024} 
  }



# Variables Used

1. max_stem_height_m 

2. average_fruit_length_cm



# Data Exploration and Cleaning
## load the necessary libraries
```{r}
library(tidyr)
library(ggbeeswarm)
library(cowplot)
library(ggridges)
library(ggplot2)
library(tidyverse)
library(skimr)
```


## import the data
```{r}
plm_data <- read.csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-18/palmtrees.csv')

str(plm_data)

skim(plm_data) 

print(plm_data)

# check for the different types of tribes we have in our dataset
unique(plm_data$palm_tribe) # 23 tribes in total

# select for species in tribes from africa and america/europe based on (Baker & Dransfield, 2016 and Eiserhardt et al., 2011) 
afro_tribe <- c("Phoeniceae", "Borasseae", "Oranieae", "Sclerospermeae", "Pelagodoxeae", "Podococceae")

eu_us_tribe <- c("Cocoseae", "Phytelepheae", "Geonomateae", "Ceroxyleae", "Chamaedoreeae", "Cryosophileae", "Trachycarpeae", "Irarteeae", "Lepidocaryeae", "Euterpeae", "Eugeissoneae", "Leopoldinieae", "Manicarieae", "Reinhardtieae", "Roystoneeae", "Sabaleae", "Cyclosphatheae")

# select for tribes from africa and take out NAs
plm_data %>%
  filter(palm_tribe %in% afro_tribe) %>%
  drop_na() -> afro_palm 

skim(afro_palm)

print(afro_palm)

# select for tribes from EU and US and take out NAs
plm_data %>% 
  filter(palm_tribe %in% eu_us_tribe) %>%
  drop_na() -> eu_us_palm 

skim(eu_us_palm)

print(eu_us_palm)

# excellent! this would be the only data wrangling we would be doing for now. 
```



# Data Visualization
## visualizing interactions for Tribes in Africa 
```{r}
# violinplot for Average Fruit Length by tribe
ggplot(afro_palm, aes(x = palm_tribe, 
                      y = average_fruit_length_cm, 
                      fill = palm_tribe)) +
  geom_violin() +
  geom_point(position=position_jitter(width=0.2,
  height=0.3),color="black",size=0.7) +
  theme_classic() +
  labs(title = "Average Fruit Length by Tribe", 
       x = "Tribe", 
       y = "Average Fruit Length (cm)")


# Boxplot for Maximum Stem Height by Tribe
ggplot(afro_palm)+ 
  aes(x = palm_tribe, 
      y = max_stem_height_m, 
      fill = palm_tribe) +
  geom_boxplot() +
  theme_cowplot() +
  labs(title = "Maximum Stem Height by Tribe", 
       x = "Tribe", 
       y = "Maximum Stem Height (m)")

#looks like our data may not be large enough for the chosen types of visualizations. I am going to use a regular boxplot and a violin plot. 
```



## visualizing interactions for Tribes in the EU and US
```{r}
# Beeswarm plot for Average Fruit Length across tribes
ggplot(eu_us_palm) +
  aes(x = palm_tribe, 
      y = average_fruit_length_cm, 
      color = palm_tribe) +
  geom_beeswarm() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12)) +
  labs(title = "Plot of Average Fruit Length Across Tribes",
       x = "Tribe",
       y = "Average Fruit Length (cm)")


# Density ridgelines for Maximum Stem Height across tribes
ggplot(eu_us_palm) +
       aes(x = max_stem_height_m, 
           y = palm_tribe, 
           fill = palm_tribe) +
  geom_density_ridges(alpha = 0.7) +
  theme_ridges() +
  theme_cowplot() +
  labs(title = "Maximum Stem Height by Tribe", 
       x = "Maximum Stem Height (m)", 
       y = "Tribe")
```

